Title Goes Here

### 2021-03-23 analysis/paper/paper.Rmd
Comentarios Lean 1ra version

# Methods

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985 (Equation 5.7). For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap].

@@

We computed the level-dependent Southern Annular Mode index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021]. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

## Complex Empirical Orthogonal Functions

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October geopotential height at 50 hPa and 100 hPa. It is clear that at 50 hPa the first two EFS represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3 pattern shifted by 1/4 wavelength. Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984]. This methods involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects.

The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

```{r corr-ceof-splitted}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)] %>%
  dcast(item1 ~ item2, value.var = "r2") %>%
  setnames("item1", "200 hPa") %>%
  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
                                            scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 3)) %>%
  kable_classic() 
```

Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EFO1 at 50 hPa. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa. Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels.

Both observations motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial part that depends on longitude, latitude and level, an a temporal part that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign. For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984]. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This was done after some exploratory analysis.

@@

@@

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimised, which also nearly maximises the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b). It's important to reassure the reader that this procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase. If the relationship is very weak, the rotation cannot generate a strong relationship.

@@

## Description of EOFs

@@

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009. Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2009.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2) show, respectively, the spatial and temporal parts of the first two leading complex EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa. In the spatial patterns in Figure\ \@ref(fig:ceofs-1), the real (in shading) and the imaginary parts (in contour) are in quadrature by construction, so that each EOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the complex temporal EOF.

The wave patterns described by these complex EOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is a wave 1, while the second is a wave 3. Note that in Figure\ \@ref(fig:ceofs-2), both parts of the EOF1 have non zero mean. This is due to th fact that the geopotential fields that enter into the algorithm are anomalies with respect to the zonal mean, not the time mean. The variability associated with the first EOF includes variability that projects onto the mean zonally anomalous field.

@@

Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950.There is an upward trend in the real part of EOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex parts of EOF2. The positive trend in the Real EOF1 translates into a positive trend in EOF1 magnitude, but not in phase (not shown).

## Regressions

The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are "pure" EOF patterns. But as they are derived by removing the zonally symmetric circulation, they might not include all the variability that is actually associated with the EOF time series. They are also idealised in that they are forced by construction, to be Hilber transforms of each other. To get a more realistic view of the real-world variability described by these EOFS, we computed regression patterns of each EOF with the complete geopotential fields.

@@

Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns of EOF1 and geopotential height. At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the Real EOF1 located towards the dateline, and the one of the Imaginary EOF1 located over Eastern Antarctica. However, the Real EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

Similarly affected are the regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b). It is only possible to distinguish some of the wave 1 pattern in relation with the Real EOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1). The Imaginary EOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the Imaginary EFO1, it's clear that these patterns are very different than the idealised versions shown in Figure\ \@ref(fig:ceofs-1), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa, suggesting that using the complex EOF method is "forcing" a wave structure where there is none.

Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere they are associated with slightly off-centre monopoles.

@@

Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the EOF2. Unlike in the case of EOF1, in this case the regression patterns represent relatively well the *idealised* patterns from Figure\ \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real EOF2 are 90º out of phase with those associated with the Imaginary EOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere.

EOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figure 1 a and b from @mo2001, the Real EOF2 can roughly be identified with the PSA2, while the Imaginary EOF2 resembles PSA1. This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability.

@@

## Relationship with other variables

### Ozone

@@

(ref:t-vertical-cap) Regression of EOF1 with zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 25ºS for the period 1979 -- 2019.

@@

Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of EOF1 with air temperature. In both levels, the distribution of temperature anomalies mirror the geopotential height anomalies in 50 hPa. To understand the vertical distribution of these anomalies, Figure\ \@ref(fig:t-vertical)) shows the regression pattern of EOF1 with zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS. Temperature zonal anomalies associated with EOF1 show a clear wave 1 pattern throughout the atmosphere above 200 hPa with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa; that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere. This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere.In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].

@@

(ref:o3-regr-cap) Regression coefficients of the real and imaginary part of the first complex EOF on SON mean Total Ozone Column. On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines). For  the 1979 -- 2019 period.

@@

Figure\ \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of EOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993]. Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone whole while the one for the Imaginary EOF1 is shifted by 90º.

@@

Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`. As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

@@

We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components .

@@

Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2. As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure\ \@ref(fig:psa-eof2)a.2). Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2). So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2. It is also the same that maximises this clean association between EFO2 parts and PSA modes. The reason that the "naive" principal component analysis arrives naturally to this optimal separation is that the imaginary phase is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.

(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019.

>>>>

# Data and Methods

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and Total Ozone COlumn (TOC) at 2.5\degree\ longitude by 2.5\degree\ latitude of horizontal resolution and 37 vertical isobaric levels from ERA5 [@era5] for the period 1979 to 2019. We restricted most of our analysis to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 [@era5be] to look at long-term trends. We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK [@fishpack] and computed horizontal wave activity fluxes following @plumb1985. For precipitation data we used monthly data from the CPC Merged Analysis of Precipitation [@cmap], with a 2.5\degree\ resolution in latitude and longitude. This rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite
inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

## Methods

\textcolor{red}{Aclarar qué a lo largo del trabajo se utilizan los datos de primavera y cómo se calcula eso. Tambień creo que acá se podría aclarar que el análisis se concentra en dos niveles para representar la estratósfera y la troposféra alta}

@@

We computed the level-dependent Southern Annular Mode (SAM) index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period [@baldwin2009]. We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively). These are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields [@campitelli2021]. Since the analysis here is only for the SON trimester, monthly values were averaged across trimesters.

@@

\textcolor{red}{Creo que habría que expandir un poco esta parte e ir comentando todoas las metodologías que vas usando: regresiones, tendencias, cómo evlauas significancia en cada caso, etc. }

## Complex Empirical Orthogonal Functions (cEOF)

(ref:eof-naive-cap) Spatial patterns of the leading 4 EOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019 (arbitrary units). The numbers at the top-left of each panel ate the zonal variance explained by each EOF.

@@

In traditional EOF analysis zonal waves appear as pairs of EOFs, usually degenerate, that represent similar patterns but shifted in phase \textcolor{red}{(Agregaría algunas referencias)}. For instance, Figure\ \@ref(fig:eof-naive) shows the leading 4 EOFs of zonally anomalous September through October \textcolor{red}{(November supongo. Sería bueno que ya arriba indiques lo de SON y ya después lo menciones directamente así)} geopotential height at 50 hPa and 200 hPa. It is clear that at 50 hPa the first two EOFs represent the same zonal wave 1 pattern and the last 2 represent the same zonal wave 3  \textcolor{red}{(No es visualmente obvia una estructura de onda 3. Solo se ven 4 centros)} pattern shifted by 1/4 wavelength. Similarly, EOFs 2 and 3 at 200 hPa represent the a zonal wave 3 pattern shifted by 1/4 wavelength. Since these pairs of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise these waves [@horel1984]. This method involves augmenting the original fields with the Hilbert transform of the data and then computing the Singular Value Decomposition of it. In this work, instead of applying the Hilbert transform to the time series at each point, we apply it to each latitude circle at each moment in time. Since the latitude circle is a periodic domain, this procedure does not suffer from edge effects. The result of the cEOF is a set of complex spatial patterns and complex time series. The real and imaginary part of each spatial pattern represent two phases wave-like spatial pattern that are in quadrature. The magnitude and argument of each complex time series represent the amplitude and phase of each zonal wave.

@@

```{r corr-ceof-splitted}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)] %>%
  dcast(item1 ~ item2, value.var = "r2") %>%
  setnames("item1", "200 hPa") %>%
  kbl2(booktabs = TRUE, caption = "Coefficient of determination ($R^2$) of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level.") %>%
  column_spec2(2:4, background = spec_color2(.table[[.col]], option = background_palette,
                                            scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 3)) %>%
  kable_classic() 
```

Another characteristic seen in Figure\ \@ref(fig:eof-naive) is that EOF1 at 200 hPa is very similar to EOF1 at 50 hPa  \textcolor{red}{(En la figura que yo veo, no diría que no similares. Aclara mejor a qué te referis)}. The zonal wave 3 structure in EOF2 and EOF3 at 200 hPa are also very similar to the one in EOF3 and EOF4 at 50 hPa, excecpt for the behaviour above the Indian Ocean. Furthermore, Table \@ref(tab:corr-ceof-splitted) shows the coefficient of determination between time series of the amplitude of each complex EOF across levels. There's a high degree of correlation between the respective EOF1 and EOF2 at each level. Both the spatial similarities and temporal correlations suggest some level of joint variability between levels. \textcolor{red}{(Revisar el parráfo considerando el comentario)}

Both observations \textcolor{red}{(¿Qué serían las dos osbervaciones?)}  motivate the decision of performing complex EOF jointly between levels. The computation of the EOFs was carried out using data from both levels at the same time, therefore, each complex EOF has a spatial component that depends on longitude, latitude and level, and a temporal component that depends only on time.

The phase of principal components is defined up to an additive constant. For real principal components, this constant can be either 0 or $\pi$, corresponding to a change in sign. For complex principal components, it can be any number between 0 and $2\pi$ [@horel1984]. Since any choice is arbitrary and equally valid, we chose the phase of each EOF so that the real and imaginary parts are aligned with meaningful phases in our analysis. This procedure does not create a spurious correlation, it only takes whatever relationship that already exist and aligns it with a specific phase.

@@

\textcolor{red}{(No mostraría la figura 2. Me parece que agrega una complejidad extra y no sé entiende demasiado, sino se la explica muy bien)}

@@

For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index [@bamston1997] and the real part was minimized, which also nearly maximizes the correlation with the imaginary part (Figure\ \@ref(fig:rotations)b).

@@

## Zonal anomalies circulation modes (no me convence del todo, pero debería ser algo así)

@@

```{r ceofs-1, fig.cap = "Spatial patterns of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019. Real part in shading, imaginary part in contours", fig.width=width_column, fig.height=width_column}
ceof[season == "SON"] %>%
  plot_ceof() +
  coord_polar() +
  axis_labs_smol +
  tag_facets(tag = "cr")
```

```{r ceofs-2, fig.cap = "Time series of the leading 2 complex EOFs of zonal anomalies of geopotential height in 50 hPa and 200 hPa for the SON trimester and the period 1979 -- 2019.", fig.height=width_column, fig.width = width_column}
ceof %>%
  .[, eof[[1]]$left, by = season] %>%
  sep_ReIm(hgt) %>%
  .[season == "SON"] %>%
  # .[, hgt := scale(hgt), by = .(part, PC)] %>%
  ggplot(aes(time, hgt)) +
  geom_hline(yintercept = 0, size = 0.1, color = "gray30") +
  geom_line(aes(color = part)) +
  scale_x_datetime(NULL) +
  scale_y_continuous("EOF") +
  scale_color_brewer(NULL, palette = "Dark2") +
  facet_grid(PC ~ .) +
  theme(panel.spacing = grid::unit(1.5, "lines"))  +
  ggthemes::geom_rangeframe(sides = "lr", size = 0.1, color = "gray30") +
  panel_background
```

To describe the spatial and temporal variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa are shown in Figures \@ref(fig:ceofs-1) and \@ref(fig:ceofs-2). The first mode (CEOF1) explains 82% of the variance, while the second mode (CEOF1) explains a smaller fraction (7%).  In the spatial patterns (Fig.\ \@ref(fig:ceofs-1)), the real (in shading) and the imaginary components (in contour) are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e. phase) is controlled by the magnitude and phase of the complex temporal EOF. The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure\ \@ref(fig:eof-naive): The first is mainly a wave 1, while the second is a wave 3.

The temporal variability for both components of the cEOF1 have non zero mean (Fig.\ \@ref(fig:ceofs-2)). This is due to the fact that the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean, not the time mean. As a consequence, the variability associated with the first cEOF includes variability that projects onto the mean zonally anomalous field.

\textcolor{red}{(Acá creo que habría que ampliar un poco más la discusión. diciendo en más detalle lo que se ve en la parte espacial (donde se ven los centros más importantes, qué diferencias hay entre estratósfera y troposféra) y temporal, por ej. qué tipo de variabilidad se observa, si están correlacionadas)}

@@

Figure\ \@ref(fig:extended-series) shows time series of the two complex EOFs extended beyond the satellite era using the preliminary ERA5 back extension going back to 1950. \textcolor{red}{(Me viene una duda metodológica que vendría bien aclarar. Cuando decis que lo extendes, ¿es que lo calculaste para todo ese período o que proyectas el modo en los otros tiempos?)} There is a significant upward trend in the real component of cEOF1 (Figure\ \@ref(fig:extended-series)a.1, p-value `r trends[part == "Real" & PC == "PC1"]$p_report`) and no significant trend in any of the complex components of cEOF2. The positive trend in the real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown). \textcolor{red}{(¿Qué interpretación se le podría dar a esto? ¿Que las amplitudes de las anomalías asociadas al modo son cada vez mayores?)}

## Regressions  (me parece que habría que decirlo de una forma más precisa o juntarlo a la sección anterior)

The spatial patterns shown in Figure\ \@ref(fig:ceofs-1) are derived by removing the zonally symmetric circulation, so they might not include all the variability that is actually associated with the cEOF time series. They are also idealised in that they are forced by construction, to be Hilbert transforms of each other \textcolor{red}{(¿Cómo se interpretaría esto?, la oración no es muy clara)}. To get a more realistic view of the real-world variability described by these cEOFs, we computed regression patterns of each cEOF with the complete geopotential fields.

\textcolor{red}{(El parráfo anterior quizás hay que trabajarlo un poco más, porque tiene algunos términos que pueden no entenderse bien o que no son los más correctos ("real-world"). También quizás esto conviene directamente contarlo cuando se describe la metodología.)}

@@

Figure\ \@ref(fig:eof1-regr-gh) shows regression patterns between geopotential height anomalies and cEOF1. At 50 hPa (Figure\ \@ref(fig:eof1-regr-gh) row a), both the Real and Imaginary EOF1 are associated with planetary wave 1 patterns, that are 90º off phase. Their phases coincide with the ones shown in Figure\ \@ref(fig:ceofs-1)a.1, with the positive centre of the real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over Eastern Antarctica. However, the cReal EOF1 pattern is substantially altered by the zonally symmetric circulation. Instead of a clear wave 1 pattern, the regression pattern can be better describes as a monopole with its centre displaced from the South Pole.

The regression patterns at 200 hPa (Figure\ \@ref(fig:eof1-regr-gh) row b) are similarly influenced by the zonally symmetric circulation. It is only possible to distinguish partially the wave 1 pattern in relation with the real cEOF1 (Figure\ \@ref(fig:eof1-regr-gh)b.1). The imaginary cEOF1 shows a much more zonally symmetrical pattern resembling a negative Southern Annular Mode.

With the exception of the imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.\ \@ref(fig:ceofs-1)), particularly at 200 hPa. Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º. This should not be utterly surprising, since the spatial patterns of the *naive* EOFs only show this feature at 50 hPa (Fig.\ \@ref(fig:eof-naive)), suggesting that using the cEOF method is artificially generating a wave structure at 200hPa. Therefore, the magnitude and phase of the EOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere. While in the troposphere, they are associated with slightly off-centre monopoles.

@@

Figure\ \@ref(fig:eof2-regr-gh) shows the regression pattern of geopotential height and the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally assymetric patterns from Figure\ \@ref(fig:ceofs-1). Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the real cEOF2 are 90º out of phase with those associated with the imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents a equivalent barotropic wave train that is very similar to the the Pacific South American Patterns [@mo2001]. Comparing the location of the positive anomaly near 90ºW in column b of Figure\ \@ref(fig:eof2-regr-gh) with Figures 1.a and b from @mo2001, the real cEOF2 can roughly be identified with PSA2, while the imaginary cEOF2 resembles PSA1. This separation is not completely meaningful in this case, since by treating this pattern as a single wave train any one pair of orthogonal waves can characterise this variability \textcolor{red}{(No me quedo del todo claro el sentido de esta última oración. Si podes ampliarla o reescribirla de otra manera, mejor)}.

@@

## cEOFs relationship with other variables

### Temperature and Ozone

@@

(ref:t-vertical-cap) Regression coefficients between cEOF1 and zonal anomalies of mean air temperature (shaded) and ozone mixing ratio (contours, negative contours with dashed lines, labels in parts per billion by mass) averaged between 80ºS and 35ºS for the period 1979 -- 2019.

@@

Figure\ \@ref(fig:eof1-regr-t) shows regression patterns of air temperature at 50hPa and 200hPa onto cEOF1. In both levels, the distribution of temperature anomalies \textcolor{red}{(Ojo que cuando se habla de regresiones uno no muestra anomalías sino coeficientes de regresión)} mirror the geopotential height anomalies in 50 hPa (Fig. XX) \textcolor{red}{(Dar un poco más de detalle de lo que se ve, aunque sea lo mismo de antes)}. The vertical distribution of these patterns is shown in Figure\ \@ref(fig:t-vertical)) that shows the regression coefficientes between cEOF1 and zonal anomalies of air temperature and zonal anomalies of ozone mixing ratio averaged between 80ºS and 35ºS. Temperature zonal anomalies associated with cEOF1 show a clear wave 1 pattern for both real and imaginary components throughout the atmosphere above 200 hPa  \textcolor{red}{(¿Empiezan en 200 hPa los contornos?)}  with a change in sign above 10 hPa. Following hydrostatic balance, this is the level in which the geopotential anomaly have maximum amplitude (not shown).

The maximum ozone anomalies are co-located with the minimum temperature anomalies above 10 hPa and with the maximum temperature anomalies below 10 hPa (Fig.\ \@ref(fig:t-vertical))); that is, the ozone zonal wave 1 is anticorrelated with the temperature zonal wave 1 in the upper stratosphere, and directly correlated in the upper stratosphere  \textcolor{red}{(¿Sabes si se le puede creer mucho a los datos allá arriba o agregamos alguna nota precautoria?)} . This change in phase is observed in ozone anomalies forced by planetary waves that reach the stratosphere. In the photochemically-dominated upper stratosphere, cold temperatures inhibit the destruction of ozone, and the advectively-dominated lower stratosphere, ozone anomalies are 90º out of phase with horizontal and vertical transport, which is 90º our of phase with temperature anomalies [@hartmann1979;@wirth1993;@smith1995].

@@

(ref:o3-regr-cap) Regression coefficients of SON mean Total Ozone Column anomalies onto the real (a) and imaginary (b) components of the cEOF1. On contours, the mean zonal anomaly of Total OZone Column (negative contours in dashed lines). ¿Units? For  the 1979 -- 2019 period.

@@

Figure\ \@ref(fig:o3-regr) shows regression maps of EOF1 with fields of Total Ozone Column (TOC). It shows zonal wave 1 pattens in TOC associated with both phases of EOF1. Climatologically, the springtime Ozone minimum is located off the South Pole and towards the Weddell Sea [@wirth1993]. Thus, the Real EOF1 regression pattern (Figure\ \@ref(fig:o3-regr)a) coincides with the climatological position of the ozone hole while the one for the Imaginary EOF1 is shifted by 90º.

@@

Indeed, Figure\ \@ref(fig:wave1-o3) shows the relationship between amplitude and phase of the planetary wave 1 in Total Ozone Column between 70ºS and 45ºS \textcolor{red}{(Habría que justifica mejor estas elecciones de bandas latitudinales a paritr del gráfico anterior)} and amplitude and phase of the EOF1. The correlation between the zonal wave 1 of ozone and EOF1 (computed as the mean cosine of the difference in phase) is `r wave_correlation$text`. As expected from the location of anomalies in Figure\ \@ref(fig:wave1-o3), the Real EOF1 drives the relationship with amplitude and the Imaginary EOF1 drives the relationship with phase (not shown).

@@

We will show that the EOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third principal components. \textcolor{red}{(Esto queda raro porque estas adelantando la conclusión)}

@@

The relationship between PSAs and the modes described from the cEOFs is also studied. Figure\ \@ref(fig:psa-eof2) shows the relationship between the two PSA indices and the Real and Imaginary phase of EOF2. As anticipated by Figure\ \@ref(fig:eof2-regr-gh), there is a strong correlation between PSA1 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2)b.1), and between PSA2 and Real EOF2 (Figure\ \@ref(fig:psa-eof2)a.2). Conversely, there is no relationship between PSA1 and Real EOF2, and between PSA2 and Imaginary EOF2 (Figure\ \@ref(fig:psa-eof2) panels a.1 and b.2). So not only this EOF2 represents well both the spatial structure and temporal evolution of the PSA modes, but it's also possible to make a rather clean association between its two phases and the two PSA modes.

This particular rotation of EOF2 is the one which maximises the relationship between ENSO and Imaginary EOF2. It is also the same that maximises this clean association between EFO2 parts and PSA modes.

\textcolor{red}{(Lo del histograma habría que explicarlo con un pocoo más detalle porque no es tan trivial. Primero decir para qué se hace, qué se hace exactamente y después recien decir cuál es el resultado y qué información aporta)}

The reason that the EOF analysis arrives to this optimal separation is that the imaginary cEOF is the most common phase (Figure\ \@ref(fig:phase-histogram)), so it naturally appears first when performing a Principal Component Analysis.

(ref:phase-histogram-cap) Histogram showing phase distribution of EOF2 for the period 1979 -- 2019. \textcolor{red}{(También ampliar)}
